---
title: "Growth Data with Intermittent Missingness"
author: "craig"
date: "2026-04-26"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
    theme: cosmo
    highlight: haddock
    code_folding: show
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '4'
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{css, echo=FALSE}
body {
  font-size: 18px;
}
pre {
  font-size: 15.6px;
  white-space: pre;
  overflow-x: visible;
}
pre code {
  white-space: pre;
  word-wrap: normal;
}
.main-container {
  max-width: 1500px !important;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "",
  fig.width = 10,
  fig.height = 6)

options(scipen = 999)
```

# Overview

This supplement presents `rblimp` R code and annotated output for the
data analysis results reported in "Missing not at Random Analyses for
Multilevel Data." The OSF repository at
[https://osf.io/yegcf/](https://osf.io/yegcf/) includes the paper and
supporting documents, and the GitHub repository at
[https://github.com/craigenders/mnar-mlm](https://github.com/craigenders/mnar-mlm)
includes executable R and Blimp Studio scripts for all examples.

The model numbers used throughout this supplement correspond to the models displayed in Tables 2 through 5 in the paper. The eight sensitivity models analyzed in this supplement are:

| Model | Description |
|:------|:------------|
| **Model 1** | CMAR |
| **Model 2** | Shared Parameter Model With Random Effects Predicting Missingness |
| **Model 3** | Shared Parameter Model With Random Effects and Their Squares Predicting Missingness |
| **Model 4** | Shared Parameter Model With Random Effect Residuals Predicting Missingness |
| **Model 5** | Selection Model With Outcome and its Lag Predicting Missingness |
| **Model 6** | Selection Model With Outcome, its Square, and its Lag Predicting Missingness |
| **Model 7** | Selection Model With Residualized Outcome and its Lag Predicting Missingness |
| **Model 8** | Model with Random Effects and Within-Person Deviations Predicting Missingness |

------------------------------------------------------------------------

# rblimp Arguments

The table below summarizes the major `rblimp` function arguments used
throughout this supplement. Each argument corresponds to a command in
the Blimp Studio scripts provided on the project's public GitHub site
([https://github.com/craigenders/mnar-mlm](https://github.com/craigenders/mnar-mlm)).
Readers unfamiliar with `rblimp` may find it helpful to use this table
as a reference while working through the analysis examples.

| Argument     | Description                                                            |
|:-------------|:-----------------------------------------------------------------------|
| `data`       | R data frame containing all variables                                  |
| `clusterid`  | Level-2 identifier variable                                            |
| `timeid`     | Integer variable indexing repeated measurement occasions               |
| `dropout`    | Creates the missingness or dropout indicator                           |
| `latent`     | Defines level-2 latent variables (random intercepts and slopes)        |
| `fixed`      | Completely observed predictor variables (speeds computation)           |
| `center`     | Specifies latent group mean centering for level-1 predictors           |
| `model`      | Specifies the focal outcome and missingness model regression equations |
| `parameters` | Defines custom derived quantities computed from model parameters       |
| `nimps`      | Number of imputations (required for computing model-predicted probabilities for graphing) |
| `seed`       | Integer random number seed for reproducibility                         |
| `chains`     | Number of independent MCMC chains (4 is also the default)             |
| `burn`       | Number of warm-up iterations per chain (discarded before estimation)   |
| `iter`       | Number of post-burn-in iterations used to compute parameter summaries  |

------------------------------------------------------------------------

# Reading the Output

The table below describes each section of the `rblimp` output in the
order it appears. Detailed descriptions of the potential scale reduction
(PSR) convergence diagnostic, the model summary tables, and the
effective sample size autocorrelation diagnostic follow the table.

| Output Section                      | Description                                                                                                                    |
|:------------------------------------|:-------------------------------------------------------------------------------------------------------------------------------|
| `BURN-IN POTENTIAL SCALE REDUCTION (PSR) OUTPUT`   | Assesses whether the MCMC warm-up period is sufficiently long; see detailed description below                                  |
| `DATA INFORMATION`                  | Gives information about the sample size and missing data rates                                                                 |
| `MODEL INFORMATION`                 | Reports the number of parameters and summarizes the fitted regression equations                                                |
| `MODEL FIT`                         | Reports DIC and WAIC and custom Wald tests, if specified                                                                       |
| `IMPUTED DATA SUMMARIES`            | Reports summary statistics for every manifest and latent variable in the model                                                 |
| `CORRELATIONS AMONG RESIDUALS`      | Reports correlations between pairs of residuals for a model's outcome variables                                                |
| `OUTCOME MODEL ESTIMATES`           | Parameter estimates, standard errors, credible intervals, and diagnostics; see detailed description below                      |
| `GENERATED PARAMETERS`              | Displays auxiliary parameters that are a function of the estimated parameters (e.g., mean difference, standardized mean difference) |

## PSR Convergence Diagnostic

The potential scale reduction factor (PSRF) or Rhat convergence
diagnostic appears near the top of the output. The PSRF assesses
whether the MCMC warm-up period is sufficiently long for the algorithm
to converge, which occurs when independent chains produce similar
parameter estimates. Smaller values closer to the theoretical minimum
of 1 are better, and values below 1.05 (Asparouhov & Muthén, 2010) or
1.10 (Gelman et al., 2014) are generally considered adequate. The
"Highest PSR" column displays the highest (worst) value across all
model parameters. This value should be less than 1.05 (or 1.10) by the
end of the burn-in period (the final row of the table). If the highest
PSRF is too large, increase the number of burn-in iterations (specified
using the `burn` argument).

```
BURN-IN RHAT CONVERGENCE DIAGNOSTIC OUTPUT:

  NOTE: Rank-normalized Rhat (Vehtari et al., 2021) is being used.
        This applies rank normalization and chain folding for robust
        convergence detection across all chains.

  Comparing iterations across 4 chains    Highest Rhat   Parameter #
                           626 to 1250           1.035            15
                          1251 to 2500           1.016            16
                          1876 to 3750           1.007             8
                          2501 to 5000           1.007            16
                          3126 to 6250           1.005            15
                          3751 to 7500           1.006             1
                          4376 to 8750           1.007            16
                          5001 to 10000          1.008            16
                          5626 to 11250          1.003            15
                          6251 to 12500          1.003            15
                          6876 to 13750          1.004            15
                          7501 to 15000          1.004            16
                          8126 to 16250          1.004            15
                          8751 to 17500          1.003             1
                          9376 to 18750          1.005             1
                         10001 to 20000          1.003             1
                         10626 to 21250          1.002            16
                         11251 to 22500          1.003            16
                         11876 to 23750          1.004            16
                         12501 to 25000          1.004             1
```

## Outcome Model Estimates

Each outcome variable has its own summary table containing raw and
standardized coefficients, variances and standard deviations, and
variance explained effect sizes. The Estimate column gives the point
estimates (posterior median), the StdDev column displays "Bayesian
standard errors" (posterior standard deviations), the 2.5% and 97.5%
columns display 95% credible intervals. The ChiSq and PValue columns
are frequentist significance tests (Wald chi-square values).

```
Latent Variable:   alpha

Parameters                         Estimate   StdDev     2.5%    97.5%       ChiSq  PValue     N_Eff
-----------------------------------------------------------------------------------------------------

Variances:
  Residual Var.                       0.330    0.055    0.227    0.442         ---     ---   1405.347

Coefficients:
  Intercept                           6.338    0.059    6.221    6.451   11671.323   0.000   5757.221
  group                               0.019    0.079   -0.134    0.178       0.059   0.809   5251.418

Standard Deviations:
  Residual SD                         0.575    0.048    0.476    0.665         ---     ---   1368.148

Standardized Coefficients:
  group                               0.017    0.069   -0.116    0.153       0.058   0.810   5223.293

Proportion Variance Explained
  by Coefficients                     0.002    0.007    0.000    0.025         ---     ---   9066.038
  by Residual Variation               0.998    0.007    0.975    1.000         ---     ---   9066.038

-----------------------------------------------------------------------------------------------------
```

The tables for alpha (random intercept) and beta (random slope)
summarize the focal model's level-2 regression equations and random
coefficient variation. The table for y displays the level-1
(within-person) residual variance.

The table for m displays the missingness model parameters. The
missingness model's pseudo-R2 (the proportion of variance explained on
the latent response metric) is often the clearest indicator of model
quality. Values around .70 or higher suggest that the predictors almost
perfectly classify dropouts from completers, a condition known as
near-separation. Very large effect sizes almost always signal that a
model lacks support from the data.

## N_Eff Diagnostic

The rightmost column in the parameter summary tables labeled N_Eff is
the effective sample size (ESS) diagnostic. The ESS provides a gauge of
whether the post-warm-up parameter summaries have sufficient precision.
Numerically, it reflects the effective number of MCMC iterations (the
"sample size") that contribute to these summaries after accounting for
autocorrelation in the sampling process. Larger values are better, and
values above 100 are generally considered adequate (Gelman et al., 2014,
p. 287). Inadequate PSRF and ESS values indicate that more iterations
are needed.

------------------------------------------------------------------------

## Setup

### Load Packages

```{r packages}
library(ggplot2)
library(patchwork)
library(rblimp)
```

### Read Data

```{r load-data}
filepath1 <- 'https://raw.githubusercontent.com/craigenders/mnar-mlm/main/growth-intermittent.csv'

growth_intermittent <- read.csv(filepath1, stringsAsFactors = T)
```

### Source Plotting Functions

```{r source-functions}
source('https://raw.githubusercontent.com/blimp-stats/blimp-book/main/misc/functions.R')
source('https://raw.githubusercontent.com/craigenders/mnar-mlm/main/mnar-plotting.R')
```

### Helper Functions

```{r helper-functions, class.source="fold-hide"}
# function to extract key parameter estimates
extract_growth_params <- function(object, method) {

  tab <- object@estimates

  rows <- c(
    "alpha ~ Intercept",
    "beta ~ Intercept",
    "alpha ~ group",
    "beta ~ group",
    "alpha residual variance",
    "beta residual variance",
    "Cor( alpha, beta )",
    "y residual variance",
    "Parameter: diff",
    "Parameter: d_diff",
    "m R2: Coefficients"
  )

  res <- do.call(rbind, lapply(rows, function(r) {
    if (r %in% rownames(tab)) {
      round(tab[r, c("Estimate", "StdDev"), drop = FALSE], 2)
    } else {
      data.frame(Estimate = NA_real_, StdDev = NA_real_, row.names = r)
    }
  }))

  rownames(res) <- c(
    "Icept (G = 0)",
    "Slope (G = 0)",
    "Icept Diff.",
    "Slope Diff.",
    "Var(Icept)",
    "Var(Slope)",
    "Cor(Icept, Slope)",
    "Var(Residual)",
    "Mean Diff.",
    "Std. Diff.",
    "Pseudo-Rsq"
  )

  colnames(res) <- c(
    paste0("Est_", method),
    paste0("SE_", method)
  )

  res
}

# function to extract convergence diagnostics
extract_convergence <- function(object, method) {

  neff <- object@estimates[, "N_Eff"]
  neff <- neff[is.finite(neff)]

  psr_row <- as.numeric(object@psr[20, ])
  psr_row <- psr_row[is.finite(psr_row)]

  data.frame(
    Min_Neff = round(min(neff),    3),
    Max_Neff = round(max(neff),    3),
    Min_PSR  = round(min(psr_row), 3),
    Max_PSR  = round(max(psr_row), 3),
    row.names = method
  )
}
```

------------------------------------------------------------------------

# Model 1: CMAR

Model 1 fits a two-level growth curve under the conditionally missing
at random (CMAR) assumption. This model serves as the baseline against
which the MNAR models are compared.

```{r model1, cache=TRUE}
model1 <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  latent = 'l2id = alpha beta',
  fixed = 'group time',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;',
  parameters = '
    diff = (((g0a+g1a)  + 4*(g0b+g1b)) - (g0a + 4*g0b));
    d_diff = diff / sqrt(y.totalvar + alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model1)
```

------------------------------------------------------------------------

# Missingness Probabilities

This section fits a series of models to examine time-related changes in missingness probabilities. The goal is to characterize how missingness rates evolve across occasions and whether those trends differ by group. Graphical and numerical summaries compare the average individual-level model-predicted probability of missingness against the observed proportions at each time point.

## ICC

The model in this section fits an unconditional two-level model for the
missingness indicator to estimate the proportion of missing data
variability attributable to stable between-person differences.

```{r icc-missingness, cache=TRUE}
icc_missingness <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  model = 'm ~ intercept | intercept;',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(icc_missingness)
```

## Linear Model

The model in this section regresses the missingness indicator on time,
group, and their interaction to assess whether missing data rates change
linearly across occasions and whether that trend differs by group.

```{r time-linear, cache=TRUE}
time_linear <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'time group',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    m ~ intercept time group time*group | intercept;',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(time_linear)
```

## Quadratic Model

The model in this section extends the linear trend by adding a
quadratic time term, allowing missingness rates to accelerate or
decelerate across occasions.

*Important features of the missingness model:*

- Squared terms are specified using `^2`.

```{r time-quadratic, cache=TRUE}
time_quadratic <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'time group',
  model = '
    level2:
    alpha ~ 1 group;
    beta ~ 1 group;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    m ~ intercept time time^2 group time*group time^2*group | intercept;',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(time_quadratic)
```

## Dummy-Coded Model

The model in this section uses dummy codes for each time point to
estimate occasion-specific missingness probabilities freely, without
imposing a functional form on how rates change over time.

*Important features of the missingness model:*

- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.

```{r time-dummy, cache=TRUE}
time_dummy <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'time group',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    m ~ intercept group | intercept;
    { t in 1:4 } : m ~ (time == [t]) (time == [t])*group;',
  parameters = '
    diff = (((g0a+g1a)  + 4*(g0b+g1b)) - (g0a + 4*g0b));
    d_diff = diff / sqrt(y.totalvar + alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(time_dummy)
```

## Missingness Probability Plots

```{r miss-plots}
ymax <- .35
ymin <- 0

plot_pmiss_obs <- plot_means(m ~ time | group,
    model = time_dummy,
    ylab = "Probability",
    title = "Observed Probabilities",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss_linear <- plot_means(m.1.probability ~ time | group,
    model = time_linear,
    ylab = "Probability",
    title = "Linear Time",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss_quadratic <- plot_means(m.1.probability ~ time | group,
    model = time_quadratic,
    ylab = "Probability",
    title = "Quadratic Time",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss_dummy <- plot_means(m.1.probability ~ time | group,
    model = time_dummy,
    ylab = "Probability",
    title = "Dummy Coded Time",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss <- (plot_pmiss_obs | plot_pmiss_linear) / (plot_pmiss_quadratic | plot_pmiss_dummy)
plot_pmiss
```

## Probability Comparison

The table below compares observed missingness probabilities to model-predicted probabilities from the linear, quadratic, and dummy-coded time models.

```{r miss-fit}
pmiss_obs       <- aggregate(m ~ time + group,               data = time_dummy@average_imp,     mean)
pmiss_linear    <- aggregate(m.1.probability ~ time + group, data = time_linear@average_imp,    mean)
pmiss_quadratic <- aggregate(m.1.probability ~ time + group, data = time_quadratic@average_imp, mean)
pmiss_dummy     <- aggregate(m.1.probability ~ time + group, data = time_dummy@average_imp,     mean)

rmse_pmiss_linear    <- sqrt(mean((pmiss_linear$m.1.probability    - pmiss_obs$m)^2))
rmse_pmiss_quadratic <- sqrt(mean((pmiss_quadratic$m.1.probability - pmiss_obs$m)^2))
rmse_pmiss_dummy     <- sqrt(mean((pmiss_dummy$m.1.probability     - pmiss_obs$m)^2))

miss_fit <- data.frame(
  RMSE   = round(c(rmse_pmiss_linear, rmse_pmiss_quadratic, rmse_pmiss_dummy), 3),
  Min    = round(c(
    min(pmiss_linear$m.1.probability    - pmiss_obs$m),
    min(pmiss_quadratic$m.1.probability - pmiss_obs$m),
    min(pmiss_dummy$m.1.probability     - pmiss_obs$m)), 3),
  Median = round(c(
    median(pmiss_linear$m.1.probability    - pmiss_obs$m),
    median(pmiss_quadratic$m.1.probability - pmiss_obs$m),
    median(pmiss_dummy$m.1.probability     - pmiss_obs$m)), 3),
  Mean   = round(c(
    mean(pmiss_linear$m.1.probability    - pmiss_obs$m),
    mean(pmiss_quadratic$m.1.probability - pmiss_obs$m),
    mean(pmiss_dummy$m.1.probability     - pmiss_obs$m)), 3),
  Max    = round(c(
    max(pmiss_linear$m.1.probability    - pmiss_obs$m),
    max(pmiss_quadratic$m.1.probability - pmiss_obs$m),
    max(pmiss_dummy$m.1.probability     - pmiss_obs$m)), 3),
  row.names = c("Linear", "Quadratic", "Dummy")
)
```

```{r miss-fit-display, echo=FALSE}
knitr::kable(miss_fit, caption = "Missingness Model Fit")
```

------------------------------------------------------------------------

# Model 2: Shared Parameter

Model 2 is a shared parameter model that links the probability of
missingness to each person's random intercept and slope, so that
individuals with higher or lower trajectories have systematically
different rates of missing data.

*Important features of the missingness model:*

- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.

```{r model2, cache=TRUE}
model2 <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'group time',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    m ~ intercept group alpha beta | intercept;
    { t in 1:4 } : m ~ (time == [t]) (time == [t])*group;',
  parameters = '
    diff = (((g0a+g1a)  + 4*(g0b+g1b)) - (g0a + 4*g0b));
    d_diff = diff / sqrt(y.totalvar + alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model2)
```

------------------------------------------------------------------------

# Model 3: Shared Parameter

Model 3 extends the shared parameter model by adding quadratic terms
for the random intercept and slope in the missingness equation,
allowing for nonlinear dependence of missingness on trajectory
characteristics.

*Important features of the missingness model:*

- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Squared terms are specified using `^2`.

```{r model3, cache=TRUE}
model3 <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'group time',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    m ~ intercept group alpha alpha^2 beta beta^2 | intercept;
    { t in 1:4 } : m ~ (time == [t]) (time == [t])*group;',
  parameters = '
    diff = (((g0a+g1a)  + 4*(g0b+g1b)) - (g0a + 4*g0b));
    d_diff = diff / sqrt(y.totalvar + alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model3)
```

------------------------------------------------------------------------

# Model 4: Shared Parameter

Model 4 reformulates the shared parameter model using random intercept
and slope residuals rather than raw latent variable values, separating
the MNAR mechanism into a component that is orthogonal to the grouping
variable.

*Important features of the missingness model:*

- Definition variables define the random intercept and slope residuals;
  these expressions are substituted into the missingness model to make
  the specification more compact.
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.

```{r model4, cache=TRUE}
model4 <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'group time',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    alpha_res = alpha - (g0a + g1a*group);
    beta_res = beta - (g0b + g1b*group);
    m ~ intercept group alpha_res beta_res | intercept;
    { t in 1:4 } : m ~ (time == [t]) (time == [t])*group;',
  parameters = '
    diff = (((g0a+g1a)  + 4*(g0b+g1b)) - (g0a + 4*g0b));
    d_diff = diff / sqrt(y.totalvar + alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model4)
```

------------------------------------------------------------------------

# Model 5: Selection

Model 5 is a Diggle-Kenward selection model that conditions the
probability of missingness on the current and lagged outcome values.

*Important features of the missingness model:*

- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Appending `.lag` to a variable's name introduces its lag into the
  model.

```{r model5, cache=TRUE}
model5 <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'group time',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    m ~ intercept group y y.lag | intercept;
    { t in 1:4 } : m ~ (time == [t]) (time == [t])*group;',
  parameters = '
    diff = (((g0a+g1a)  + 4*(g0b+g1b)) - (g0a + 4*g0b));
    d_diff = diff / sqrt(y.totalvar + alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model5)
```

------------------------------------------------------------------------

# Model 6: Selection

Model 6 extends the Diggle-Kenward model by adding a quadratic term for
the current outcome, allowing for nonlinear dependence of missingness on
outcome values.

*Important features of the missingness model:*

- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Appending `.lag` to a variable's name introduces its lag into the
  model.
- Squared terms are specified using `^2`.

```{r model6, cache=TRUE}
model6 <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'group time',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    m ~ intercept group y y^2 y.lag | intercept;
    { t in 1:4 } : m ~ (time == [t]) (time == [t])*group;',
  parameters = '
    diff = (((g0a+g1a)  + 4*(g0b+g1b)) - (g0a + 4*g0b));
    d_diff = diff / sqrt(y.totalvar + alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model6)
```

------------------------------------------------------------------------

# Model 7: Selection

Model 7 reformulates the Diggle-Kenward model using within-person
residuals (detrended outcomes with the individual trajectory removed)
rather than raw outcome values in the missingness equation. This model
uses `burn = 200000` and `iter = 200000` to achieve adequate
convergence.

*Important features of the missingness model:*

- Definition variables define the detrended outcome variables
  (within-person deviations); these expressions are substituted into
  the missingness model to make the specification more compact.
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Appending `.lag` to a variable's name introduces its lag into the
  model.

```{r model7, cache=TRUE}
model7 <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'group time',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    yw = y - (alpha + beta*time);
    lagyw = y.lag - (alpha + beta*(time - 1));
    m ~ intercept group yw lagyw | intercept;
    { t in 1:4 } : m ~ (time == [t]) (time == [t])*group;',
  parameters = '
    diff = (((g0a+g1a)  + 4*(g0b+g1b)) - (g0a + 4*g0b));
    d_diff = diff / sqrt(y.totalvar + alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 200000,
  iter = 200000)

output(model7)
```

------------------------------------------------------------------------

# Model 8: Combined

Model 8 is a disaggregated model that combines elements of both
frameworks, conditioning missingness on trajectory characteristics
(intercept and slope) and within-person residuals simultaneously. This
model uses `burn = 200000` and `iter = 200000` to achieve adequate
convergence.

*Important features of the missingness model:*

- Definition variables define the detrended outcome variables
  (within-person deviations); these expressions are substituted into
  the missingness model to make the specification more compact.
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Appending `.lag` to a variable's name introduces its lag into the
  model.

```{r model8, cache=TRUE}
model8 <- rblimp(
  data = growth_intermittent,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (missing)',
  latent = 'l2id = alpha beta',
  fixed = 'group time',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    alpha ~~ beta;
    level1:
    y ~ intercept@alpha time@beta;
    missingness:
    yw = y - (alpha + beta*time);
    lagyw = y.lag - (alpha + beta*(time - 1));
    m ~ intercept group yw lagyw alpha beta | intercept;
    { t in 1:4 } : m ~ (time == [t]) (time == [t])*group;',
  parameters = '
    diff = (((g0a+g1a)  + 4*(g0b+g1b)) - (g0a + 4*g0b));
    d_diff = diff / sqrt(y.totalvar + alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 200000,
  iter = 200000)

output(model8)
```

------------------------------------------------------------------------

# Sensitivity Summary

This section extracts key information from the `rblimp` model objects and summarizes it in tabular form, making it easier to compare results across all sensitivity models at a glance.

## Model Key

The model numbers used throughout this supplement correspond to the models displayed in Tables 2 through 5 in the paper.

| Model | Description |
|:------|:------------|
| **Model 1** | CMAR |
| **Model 2** | Shared Parameter Model With Random Effects Predicting Missingness |
| **Model 3** | Shared Parameter Model With Random Effects and Their Squares Predicting Missingness |
| **Model 4** | Shared Parameter Model With Random Effect Residuals Predicting Missingness |
| **Model 5** | Selection Model With Outcome and its Lag Predicting Missingness |
| **Model 6** | Selection Model With Outcome, its Square, and its Lag Predicting Missingness |
| **Model 7** | Selection Model With Residualized Outcome and its Lag Predicting Missingness |
| **Model 8** | Model with Random Effects and Within-Person Deviations Predicting Missingness |

## Parameter Estimates

The table displays the estimates (posterior medians) and Bayesian standard errors (posterior standard deviations) for each parameter by model. Column headers beginning with `Est` contain the point estimates and those beginning with `SE` contain the Bayesian standard errors.

```{r extract-params, cache=TRUE, dependson=c("helper-functions","model1","model2","model3","model4","model5","model6","model7","model8"), class.source="fold-hide"}
table_summary <- cbind(
  extract_growth_params(model1, "Mod1"),
  extract_growth_params(model2, "Mod2"),
  extract_growth_params(model3, "Mod3"),
  extract_growth_params(model4, "Mod4"),
  extract_growth_params(model5, "Mod5"),
  extract_growth_params(model6, "Mod6"),
  extract_growth_params(model7, "Mod7"),
  extract_growth_params(model8, "Mod8")
)
```

```{r table-summary-display, echo=FALSE, dependson="extract-params"}
knitr::kable(table_summary, caption = "Parameter Estimates")
```

## Changes in SE Units

The table displays the differences between the MNAR model estimates and the CMAR estimates (Model 1), divided by the CMAR standard errors. Values near zero indicate that the MNAR assumption has little practical impact on that parameter relative to sampling variability. SE unit changes gauge how a test statistic reacts to different MNAR assumptions. For example, an SE unit change of 0.5 means a test statistic could shift from 1.5 to 2.0, or vice versa, depending on the direction of the change.

```{r table-change, cache=TRUE, dependson="extract-params", class.source="fold-hide"}
est_cmar <- table_summary[, "Est_Mod1"]
se_cmar  <- table_summary[, "SE_Mod1"]

compare_methods <- c("Mod2", "Mod3", "Mod4", "Mod5", "Mod6", "Mod7", "Mod8")

table_change <- sapply(compare_methods, function(m) {
  round((table_summary[, paste0("Est_", m)] - est_cmar) / se_cmar, 2)
})
table_change <- as.data.frame(table_change)
rownames(table_change) <- rownames(table_summary)
```

```{r table-change-display, echo=FALSE, dependson="table-change"}
knitr::kable(table_change, caption = "Changes vs. CMAR in Standard Error Units")
```

## Convergence Diagnostics

The table displays the minimum and maximum PSR and N_Eff values across all parameters by model. For the PSR, smaller values closer to the theoretical minimum of 1 are better, and values below 1.05 (Asparouhov & Muthén, 2010) or 1.10 (Gelman et al., 2014) are generally considered adequate. For the ESS, larger values are better, and values above 100 are generally considered adequate (Gelman et al., 2014, p. 287). Inadequate PSRF and ESS values indicate that more iterations are needed.

```{r extract-conv, cache=TRUE, dependson=c("model1","model2","model3","model4","model5","model6","model7","model8"), class.source="fold-hide"}
table_diag <- rbind(
  extract_convergence(model1, "Model 1"),
  extract_convergence(model2, "Model 2"),
  extract_convergence(model3, "Model 3"),
  extract_convergence(model4, "Model 4"),
  extract_convergence(model5, "Model 5"),
  extract_convergence(model6, "Model 6"),
  extract_convergence(model7, "Model 7"),
  extract_convergence(model8, "Model 8")
)

table_diag$Iterations <- c(
  nrow(model1@iterations),
  nrow(model2@iterations),
  nrow(model3@iterations),
  nrow(model4@iterations),
  nrow(model5@iterations),
  nrow(model6@iterations),
  nrow(model7@iterations),
  nrow(model8@iterations)
)
```

```{r table-diag-display, echo=FALSE, dependson="extract-conv"}
knitr::kable(table_diag, caption = "Convergence Diagnostics")
```

------------------------------------------------------------------------
