---
title: "Intensive Data with Dropout"
author: "craig"
date: "2026-04-26"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
    theme: cosmo
    highlight: haddock
    code_folding: show
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '4'
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{css, echo=FALSE}
body {
  font-size: 18px;
}
pre {
  font-size: 15.6px;
  white-space: pre;
  overflow-x: visible;
}
pre code {
  white-space: pre;
  word-wrap: normal;
}
.main-container {
  max-width: 1500px !important;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "",
  fig.width = 10,
  fig.height = 6)

options(scipen = 999)
```

This supplement presents `rblimp` R code and annotated output for the
data analysis results reported in "Missing not at Random Analyses for
Multilevel Data." The OSF repository at
[https://osf.io/yegcf/](https://osf.io/yegcf/) includes the paper and
supporting documents, and the GitHub repository at
[https://github.com/craigenders/mnar-mlm](https://github.com/craigenders/mnar-mlm)
includes executable R and Blimp Studio scripts for all examples.

------------------------------------------------------------------------

## rblimp Arguments

The table below summarizes the major `rblimp` function arguments used
throughout this supplement. Each argument corresponds to a command in
the Blimp Studio scripts provided on the project's public GitHub site
([https://github.com/craigenders/mnar-mlm](https://github.com/craigenders/mnar-mlm)).
Readers unfamiliar with `rblimp` may find it helpful to use this table
as a reference while working through the analysis examples.

| Argument     | Description                                                            |
|:-------------|:-----------------------------------------------------------------------|
| `data`       | R data frame containing all variables                                  |
| `clusterid`  | Level-2 identifier variable                                            |
| `timeid`     | Integer variable indexing repeated measurement occasions               |
| `dropout`    | Creates the missingness or dropout indicator                           |
| `latent`     | Defines level-2 latent variables (random intercepts and slopes)        |
| `fixed`      | Completely observed predictor variables (speeds computation)           |
| `center`     | Specifies latent group mean centering for level-1 predictors           |
| `model`      | Specifies the focal outcome and missingness model regression equations |
| `parameters` | Defines custom derived quantities computed from model parameters       |
| `nimps`      | Number of imputations (required for computing model-predicted probabilities for graphing) |
| `seed`       | Integer random number seed for reproducibility                         |
| `chains`     | Number of independent MCMC chains (4 is also the default)             |
| `burn`       | Number of warm-up iterations per chain (discarded before estimation)   |
| `iter`       | Number of post-burn-in iterations used to compute parameter summaries  |

------------------------------------------------------------------------

## Reading the Output

The table below describes each section of the `rblimp` output in the
order it appears. Detailed descriptions of the BURN-IN RHAT and OUTCOME
MODEL ESTIMATES sections follow the table.

| Output Section                      | Description                                                                                                                    |
|:------------------------------------|:-------------------------------------------------------------------------------------------------------------------------------|
| `BURN-IN RHAT CONVERGENCE OUTPUT`   | Assesses whether the MCMC warm-up period is sufficiently long; see detailed description below                                  |
| `DATA INFORMATION`                  | Gives information about the sample size and missing data rates                                                                 |
| `MODEL INFORMATION`                 | Reports the number of parameters and summarizes the fitted regression equations                                                |
| `MODEL FIT`                         | Reports DIC and WAIC and custom Wald tests, if specified                                                                       |
| `IMPUTED DATA SUMMARIES`            | Reports summary statistics for every manifest and latent variable in the model                                                 |
| `CORRELATIONS AMONG RESIDUALS`      | Reports correlations between pairs of residuals for a model's outcome variables                                                |
| `OUTCOME MODEL ESTIMATES`           | Parameter estimates, standard errors, credible intervals, and diagnostics; see detailed description below                      |
| `GENERATED PARAMETERS`              | Displays auxiliary parameters that are a function of the estimated parameters (e.g., mean difference, standardized mean difference) |

### BURN-IN RHAT Convergence Diagnostic

The potential scale reduction factor (PSRF) or Rhat convergence
diagnostic appears near the top of the output. The PSRF assesses
whether the MCMC warm-up period is sufficiently long for the algorithm
to converge, which occurs when independent chains produce similar
parameter estimates. Smaller values closer to the theoretical minimum
of 1 are better, and values below 1.05 (Asparouhov & Muthén, 2010) or
1.10 (Gelman et al., 2014) are generally considered adequate. The
"Highest PSR" column displays the highest (worst) value across all
model parameters. This value should be less than 1.05 (or 1.10) by the
end of the burn-in period (the final row of the table). If the highest
PSRF is too large, increase the number of burn-in iterations (specified
using the `burn` argument).

### Outcome Model Estimates

Each outcome variable has its own summary table containing raw and
standardized coefficients, variances and standard deviations, and
variance explained effect sizes. The Estimate column gives the point
estimates (posterior median), the StdDev column displays "Bayesian
standard errors" (posterior standard deviations), the 2.5% and 97.5%
columns display 95% credible intervals. The ChiSq and PValue columns
are frequentist significance tests (Wald chi-square values).

The rightmost column labeled N_Eff is the effective sample size (ESS)
diagnostic. The ESS provides a gauge of whether the post-warm-up
parameter summaries have sufficient precision. Numerically, it reflects
the effective number of MCMC iterations (the "sample size") that
contribute to these summaries after accounting for autocorrelation in
the sampling process. Values above 100 are generally considered
adequate (Gelman et al., 2014, p. 287). Inadequate PSRF and ESS values
indicate that more iterations are needed.

The tables for alpha (random intercept), beta (random slope), and omega
(random within-person log-variation) summarize the focal model's
level-2 regression equations and random coefficient variation. The
table for y displays the level-1 (within-person) residual variance.

The table for m displays the missingness model parameters. The
missingness model's pseudo-R2 (the proportion of variance explained on
the latent response metric) is often the clearest indicator of model
quality. Values around .70 or higher suggest that the predictors almost
perfectly classify dropouts from completers, a condition known as
near-separation. Very large effect sizes almost always signal that a
model lacks support from the data.

------------------------------------------------------------------------

## Setup

### Load Packages

```{r packages}
library(ggplot2)
library(patchwork)
library(rblimp)
```

### Read Data

```{r load-data}
filepath <- 'https://raw.githubusercontent.com/craigenders/mnar-mlm/main/intensive-dropout.csv'

intensive_dropout <- read.csv(filepath, stringsAsFactors = T)
```

### Source Plotting Functions

```{r source-functions}
source('https://raw.githubusercontent.com/blimp-stats/blimp-book/main/misc/functions.R')
source('https://raw.githubusercontent.com/craigenders/mnar-mlm/main/mnar-plotting.R')
```

### Helper Functions

```{r helper-functions}
# function to extract key parameter estimates
extract_intensive_params <- function(object, method) {

  tab <- object@estimates

  rows <- c(
    "alpha ~ Intercept",
    "beta ~ Intercept",
    "alpha ~ group",
    "beta ~ group",
    "alpha residual variance",
    "beta residual variance",
    "Cor( alpha, beta )",
    "y Variance Q50%",
    "omega ~ Intercept",
    "omega residual variance",
    "Cor( alpha, omega )",
    "Cor( beta, omega )",
    "Parameter: diff",
    "Parameter: d_diff",
    "m R2: Coefficients"
  )

  res <- do.call(rbind, lapply(rows, function(r) {
    if (r %in% rownames(tab)) {
      round(tab[r, c("Estimate", "StdDev"), drop = FALSE], 2)
    } else {
      data.frame(Estimate = NA_real_, StdDev = NA_real_, row.names = r)
    }
  }))

  rownames(res) <- c(
    "Icept (G = 0)",
    "Slope (G = 0)",
    "Icept Diff.",
    "Slope Diff.",
    "Var(Icept)",
    "Var(Slope)",
    "Cor(Icept, Slope)",
    "Var(Residual)",
    "Icept Log-Var",
    "Var(Log-Var)",
    "Cor(Icept, Log-Var)",
    "Cor(Log-Var, Slope)",
    "Mean Diff.",
    "Std. Diff.",
    "Pseudo-Rsq"
  )

  colnames(res) <- c(
    paste0("Est_", method),
    paste0("SD_", method)
  )

  res
}

# function to extract convergence diagnostics
extract_convergence <- function(object, method) {

  neff <- object@estimates[, "N_Eff"]
  neff <- neff[is.finite(neff)]

  psr_row <- as.numeric(object@psr[20, ])
  psr_row <- psr_row[is.finite(psr_row)]

  data.frame(
    Min_Neff = round(min(neff),    3),
    Max_Neff = round(max(neff),    3),
    Min_PSR  = round(min(psr_row), 3),
    Max_PSR  = round(max(psr_row), 3),
    row.names = method
  )
}
```

------------------------------------------------------------------------

# Model 1: CMAR

Model 1 fits a two-level growth curve under the conditionally missing
at random (CMAR) assumption. This model serves as the baseline against
which the MNAR models are compared.

```{r model1, cache=TRUE}
model1 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  latent = 'l2id = alpha beta omega',
  fixed = 'group time',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model1)
```

------------------------------------------------------------------------

# Missingness Probabilities

## Linear Missingness

The model in this section regresses the missingness indicator on time,
group, and their interaction to assess whether missing data rates change
linearly across occasions and whether that trend differs by group.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).

```{r time-linear, cache=TRUE}
time_linear <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'time group',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 d group*d (time - 7)*d (time - 7)*group*d | intercept@0;',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 20000,
  nimps = 20)

output(time_linear)
```

## Quadratic Missingness

The model in this section extends the linear trend by adding a
quadratic time term, allowing missingness rates to accelerate or
decelerate across occasions.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Squared terms are specified using `^2`.

```{r time-quadratic, cache=TRUE}
time_quadratic <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'time group',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 d group*d (time - 7)*d (time - 7)^2*d (time - 7)*group*d (time - 7)^2*group*d | intercept@0;',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 20000,
  nimps = 20)

output(time_quadratic)
```

## Dummy-Coded Missingness

The model in this section uses dummy codes for each time point to
estimate occasion-specific missingness probabilities freely, without
imposing a functional form on how rates change over time.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.

```{r time-dummy, cache=TRUE}
time_dummy <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'time group',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 | intercept@0;
    { t in 1:19 } : m ~ (time == [t])*d (time == [t])*group*d;',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(time_dummy)
```

## Missingness Probability Plots

```{r miss-plots}
ymax <- .15
ymin <- 0

plot_pmiss_obs <- plot_means(m ~ time | group,
    model = time_dummy,
    ylab = "Probability",
    title = "Observed Probabilities",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss_linear <- plot_means(m.1.probability ~ time | group,
    model = time_linear,
    ylab = "Probability",
    title = "Linear Time",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss_quadratic <- plot_means(m.1.probability ~ time | group,
    model = time_quadratic,
    ylab = "Probability",
    title = "Quadratic Time",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss_dummy <- plot_means(m.1.probability ~ time | group,
    model = time_dummy,
    ylab = "Probability",
    title = "Dummy Coded Time",
    group_labels = c("0" = "0", "1" = "1")) + ylim(ymin, ymax) +
    theme(legend.position = "top", legend.justification = "center") +
    scale_linetype_manual(values = c("dashed", "solid")) +
    geom_line(linewidth = .25)

plot_pmiss <- (plot_pmiss_obs | plot_pmiss_linear) / (plot_pmiss_quadratic | plot_pmiss_dummy)
plot_pmiss
```

## Probability Comparison

```{r miss-fit}
pmiss_obs       <- aggregate(m ~ time + group,               data = time_dummy@average_imp,     mean)
pmiss_linear    <- aggregate(m.1.probability ~ time + group, data = time_linear@average_imp,    mean)
pmiss_quadratic <- aggregate(m.1.probability ~ time + group, data = time_quadratic@average_imp, mean)
pmiss_dummy     <- aggregate(m.1.probability ~ time + group, data = time_dummy@average_imp,     mean)

rmse_pmiss_linear    <- sqrt(mean((pmiss_linear$m.1.probability    - pmiss_obs$m)^2))
rmse_pmiss_quadratic <- sqrt(mean((pmiss_quadratic$m.1.probability - pmiss_obs$m)^2))
rmse_pmiss_dummy     <- sqrt(mean((pmiss_dummy$m.1.probability     - pmiss_obs$m)^2))

miss_fit <- data.frame(
  RMSE   = round(c(rmse_pmiss_linear, rmse_pmiss_quadratic, rmse_pmiss_dummy), 3),
  Min    = round(c(
    min(pmiss_linear$m.1.probability    - pmiss_obs$m),
    min(pmiss_quadratic$m.1.probability - pmiss_obs$m),
    min(pmiss_dummy$m.1.probability     - pmiss_obs$m)), 3),
  Median = round(c(
    median(pmiss_linear$m.1.probability    - pmiss_obs$m),
    median(pmiss_quadratic$m.1.probability - pmiss_obs$m),
    median(pmiss_dummy$m.1.probability     - pmiss_obs$m)), 3),
  Mean   = round(c(
    mean(pmiss_linear$m.1.probability    - pmiss_obs$m),
    mean(pmiss_quadratic$m.1.probability - pmiss_obs$m),
    mean(pmiss_dummy$m.1.probability     - pmiss_obs$m)), 3),
  Max    = round(c(
    max(pmiss_linear$m.1.probability    - pmiss_obs$m),
    max(pmiss_quadratic$m.1.probability - pmiss_obs$m),
    max(pmiss_dummy$m.1.probability     - pmiss_obs$m)), 3),
  row.names = c("Linear", "Quadratic", "Dummy")
)
```

```{r miss-fit-display, echo=FALSE}
knitr::kable(miss_fit, caption = "Missingness Model Fit")
```

------------------------------------------------------------------------

# Model 2: Shared Parameter

Model 2 is a shared parameter model that links the probability of
missingness to each person's random intercept and slope, so that
individuals with higher or lower trajectories have systematically
different rates of missing data. Because the first seven occasions are
complete, the missingness model uses interactions with a time-varying
dummy code to activate effects at occasions with missing data. This
model uses `burn = 75000` and `iter = 75000` to achieve adequate
convergence.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.

```{r model2, cache=TRUE}
model2 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'group time',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 alpha*d omega*d | intercept@0;
    { t in 1:19 } : m ~ (time == [t])*d (time == [t])*group*d;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 75000,
  iter = 75000)

output(model2)
```

------------------------------------------------------------------------

# Model 3: Shared Parameter

Model 3 extends the shared parameter model by adding quadratic terms
for the random intercept and slope in the missingness equation,
allowing for nonlinear dependence of missingness on trajectory
characteristics. Because the first seven occasions are complete, the
missingness model uses interactions with a time-varying dummy code to
activate effects at occasions with missing data. This model uses
`burn = 200000` and `iter = 200000` to achieve adequate convergence.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Squared terms are specified using `^2`.

```{r model3, cache=TRUE}
model3 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'group time',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 alpha*d omega*d alpha^2*d omega^2*d | intercept@0;
    { t in 1:19 } : m ~ (time == [t])*d (time == [t])*group*d;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 200000,
  iter = 200000)

output(model3)
```

------------------------------------------------------------------------

# Model 4: Shared Parameter

Model 4 reformulates the shared parameter model using random intercept
and slope residuals rather than raw latent variable values, separating
the MNAR mechanism into a component that is orthogonal to the grouping
variable. Because the first seven occasions are complete, the
missingness model uses interactions with a time-varying dummy code to
activate effects at occasions with missing data.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data.
- Definition variables also define the random effect residuals; these
  expressions are substituted into the missingness model to make the
  specification more compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.

```{r model4, cache=TRUE}
model4 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'group time',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    alpha_res = alpha - (g0a + g1a*group);
    m ~ intercept@-3 d*alpha_res d*omega | intercept@0;
    { t in 1:19 } : m ~ d*(time == [t]) d*(time == [t])*group;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model4)
```

------------------------------------------------------------------------

# Model 5: Shared Parameter

Model 5 extends the shared parameter model by adding level-2 latent
group means for the predictor, allowing missingness to depend on both
the outcome and the covariate. Because the first seven occasions are
complete, the missingness model uses interactions with a time-varying
dummy code to activate effects at occasions with missing data. This
model uses `burn = 200000` and `iter = 200000` to achieve adequate
convergence.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Level-2 latent group means are added by appending `.mean` to the
  covariate's name.

```{r model5, cache=TRUE}
model5 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'group time',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 d*alpha d*omega d*x.mean | intercept@0;
    { t in 1:19 } : m ~ d*(time == [t]) d*(time == [t])*group;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 200000,
  iter = 200000)

output(model5)
```

------------------------------------------------------------------------

# Model 6: Selection Model

Model 6 is a Diggle-Kenward selection model that conditions the
probability of missingness on the current and lagged outcome values.
Because the first seven occasions are complete, the missingness model
uses interactions with a time-varying dummy code to activate effects at
occasions with missing data.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Appending `.lag` to a variable's name introduces its lag into the
  model.

```{r model6, cache=TRUE}
model6 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'group time',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 y*d y.lag*d | intercept@0;
    { t in 1:19 } : m ~ (time == [t])*d (time == [t])*group*d;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model6)
```

------------------------------------------------------------------------

# Model 7: Selection Model

Model 7 extends the Diggle-Kenward model by adding a quadratic term for
the current outcome, allowing for nonlinear dependence of missingness on
outcome values. Because the first seven occasions are complete, the
missingness model uses interactions with a time-varying dummy code to
activate effects at occasions with missing data. This model uses
`burn = 200000` and `iter = 200000` to achieve adequate convergence.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Appending `.lag` to a variable's name introduces its lag into the
  model.
- Squared terms are specified using `^2`.

```{r model7, cache=TRUE}
model7 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'group time',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 y*d y^2*d y.lag*d | intercept@0;
    { t in 1:19 } : m ~ (time == [t])*d (time == [t])*group*d;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 200000,
  iter = 200000)

output(model7)
```

------------------------------------------------------------------------

# Model 8: Selection Model

Model 8 reformulates the Diggle-Kenward model using within-person
residuals (outcomes centered at the latent person means) rather than
raw outcome values in the missingness equation. Because the first seven
occasions are complete, the missingness model uses interactions with a
time-varying dummy code to activate effects at occasions with missing
data. This model uses `burn = 75000` and `iter = 75000` to achieve
adequate convergence.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data.
- Definition variables also define the centered outcome variable
  (within-person deviations); these expressions are substituted into
  the missingness model to make the specification more compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Appending `.lag` to a variable's name introduces its lag into the
  model.

```{r model8, cache=TRUE}
model8 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'group time',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 (y - alpha)*d (y.lag - alpha)*d | intercept@0;
    { t in 1:19 } : m ~ (time == [t])*d (time == [t])*group*d;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 75000,
  iter = 75000)

output(model8)
```

------------------------------------------------------------------------

# Model 9: Selection Model

Model 9 extends the Diggle-Kenward model by adding the focal model's
covariate (X) and its lag as predictors, allowing missingness to depend
on both the outcome and the covariate. Because the first seven occasions
are complete, the missingness model uses interactions with a
time-varying dummy code to activate effects at occasions with missing
data.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Appending `.lag` to a variable's name introduces its lag into the
  model.
- In order to use the `.lag` function, the covariate (X) requires its
  own empty model and explicit latent group means (random intercepts).

```{r model9, cache=TRUE}
model9 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega xmeans',
  fixed = 'group time',
  model = '
    level2:
    xmeans ~ intercept;
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    x ~ intercept@xmeans;
    y ~ intercept@alpha (x - xmeans)@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 y*d y.lag*d x*d x.lag*d | intercept@0;
    { t in 1:19 } : m ~ (time == [t])*d (time == [t])*group*d;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 25000,
  iter = 25000)

output(model9)
```

------------------------------------------------------------------------

# Model 10: Combined Model

Model 10 is a disaggregated model that combines elements of both
frameworks, conditioning missingness on trajectory characteristics
(intercept and slope) and within-person residuals simultaneously.
Because the first seven occasions are complete, the missingness model
uses interactions with a time-varying dummy code to activate effects at
occasions with missing data. This model uses `burn = 200000` and
`iter = 200000` to achieve adequate convergence.

*Important features of the missingness model:*

- A definition variable creates the time-varying dummy code that
  activates effects at occasions with missing data; this expression is
  substituted into the missingness model to make the specification more
  compact.
- The `@` symbol fixes parameters at specific numeric values (e.g., -3
  for the fixed intercept and 0 for the random intercept variance after
  the vertical pipe).
- Boolean statements create dummy codes as functions of the TIME
  variable, such that each condition in parentheses returns 1 if
  TIME = `t` and 0 otherwise.
- A looping structure avoids specifying a separate Boolean statement for
  each occasion.
- Appending `.lag` to a variable's name introduces its lag into the
  model.
- Centered outcomes are created by subtracting the random intercepts
  from the outcome and its lag.

```{r model10, cache=TRUE}
model10 <- rblimp(
  data = intensive_dropout,
  clusterid = 'l2id',
  timeid = 'time',
  dropout = 'm = y (monotone)',
  latent = 'l2id = alpha beta omega',
  fixed = 'group time',
  center = 'groupmean = x',
  model = '
    level2:
    alpha ~ intercept@g0a group@g1a;
    beta ~ intercept@g0b group@g1b;
    omega ~ intercept@g0o;
    alpha beta omega ~~ alpha beta omega;
    level1:
    y ~ intercept@alpha x@beta;
    var(y) ~ intercept@omega;
    missingness:
    d = ifelse(time < 7, 0, 1);
    m ~ intercept@-3 (y - alpha)*d (y.lag - alpha)*d alpha*d omega*d | intercept@0;
    { t in 1:19 } : m ~ (time == [t])*d (time == [t])*group*d;',
  parameters = '
    diff = g1a;
    d_diff = diff / sqrt(alpha.totalvar);',
  seed = 90291,
  chains = 4,
  burn = 200000,
  iter = 200000)

output(model10)
```

------------------------------------------------------------------------

# Summary Tables

## Parameter Estimates

```{r extract-params, cache=TRUE, dependson=c("model1","model2","model3","model4","model5","model6","model7","model8","model9","model10")}
table_summary <- cbind(
  extract_intensive_params(model1,  "M1"),
  extract_intensive_params(model2,  "M2"),
  extract_intensive_params(model3,  "M3"),
  extract_intensive_params(model4,  "M4"),
  extract_intensive_params(model5,  "M5"),
  extract_intensive_params(model6,  "M6"),
  extract_intensive_params(model7,  "M7"),
  extract_intensive_params(model8,  "M8"),
  extract_intensive_params(model9,  "M9"),
  extract_intensive_params(model10, "M10")
)
```

```{r table-summary-display, echo=FALSE, dependson="extract-params"}
knitr::kable(table_summary, caption = "Parameter Estimates")
```

## Effect Sizes

```{r extract-diff, cache=TRUE, dependson="extract-params"}
mean_row <- table_summary["Mean Diff.", ]
std_row  <- table_summary["Std. Diff.", ]

methods_short <- paste0("M", 1:10)
methods_long  <- paste0("Model ", 1:10)

table_diff <- do.call(rbind, lapply(seq_along(methods_short), function(i) {
  m <- methods_short[i]
  c(
    Mean_Diff     = unname(mean_row[paste0("Est_", m)]),
    SD_Mean_Diff  = unname(mean_row[paste0("SD_",  m)]),
    Std_Mean_Diff = unname(std_row[paste0("Est_",  m)]),
    SD_Std_Mean   = unname(std_row[paste0("SD_",   m)])
  )
}))
rownames(table_diff) <- methods_long
table_diff <- as.data.frame(table_diff)
```

```{r table-diff-display, echo=FALSE, dependson="extract-diff"}
knitr::kable(table_diff, caption = "Effect Sizes")
```

## Convergence Diagnostics

```{r extract-conv, cache=TRUE, dependson=c("model1","model2","model3","model4","model5","model6","model7","model8","model9","model10")}
table_diag <- rbind(
  extract_convergence(model1,  "Model 1"),
  extract_convergence(model2,  "Model 2"),
  extract_convergence(model3,  "Model 3"),
  extract_convergence(model4,  "Model 4"),
  extract_convergence(model5,  "Model 5"),
  extract_convergence(model6,  "Model 6"),
  extract_convergence(model7,  "Model 7"),
  extract_convergence(model8,  "Model 8"),
  extract_convergence(model9,  "Model 9"),
  extract_convergence(model10, "Model 10")
)

table_diag$Iterations <- c(
  nrow(model1@iterations),
  nrow(model2@iterations),
  nrow(model3@iterations),
  nrow(model4@iterations),
  nrow(model5@iterations),
  nrow(model6@iterations),
  nrow(model7@iterations),
  nrow(model8@iterations),
  nrow(model9@iterations),
  nrow(model10@iterations)
)
```

```{r table-diag-display, echo=FALSE, dependson="extract-conv"}
knitr::kable(table_diag, caption = "Convergence Diagnostics")
```

------------------------------------------------------------------------
